Pervasive duplication of tumor suppressors in Afrotherians preceded evolution of large bodies and reduced cancer risk in elephants

### 2020-09-08 paper_PLOS/paper_PLOS_draft.tex
Penultimate draft

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Pervasive duplication of tumor suppressor genes preceded parallel evolution of large bodied Atlantogenatans},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

\setcounter{secnumdepth}{0}
% Pandoc toggle for numbering sections (defaults to be off)
\setcounter{secnumdepth}{0}

% Pandoc header
\usepackage{longtable}
\usepackage{booktabs}
\usepackage{geometry}
\usepackage{pdflscape}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

@@

  \title{Pervasive duplication of tumor suppressor genes preceded parallel
evolution of large bodied Atlantogenatans}
    \author[University of Chicago]{Juan Manuel Vazquez}

@@

  \begin{abstract}
  Cancer is an intrinsic disease of multicellular organisms. Within a
  species, the size of an animal, - correlated with the individual's
  number of cells - and its lifespan - correlated with increasing cellular
  damage over time - positively correlate with the risk any individual has
  to form tumors. Between species, however, we do not observe any
  correlation between size, lifespan, and cancer, a phenomenon that
  referred to as \emph{Peto's Paradox}. Elephants are a particularly
  interesting member of this class of paradoxical animals, since they are
  a set of large species deeply nested in a clade of smaller species,
  indicating a recent gain of size. Recent work has identified several
  individual cases of gene duplicates contributing to the increased cancer
  resistance of elephants, which suggests that duplication of tumor
  suppressor genes may play a more general role in mediating Peto's
  Paradox by increasing cancer resistance in large, long-lived species. By
  using a Reciprocal Best-Hit BLAT search approach, we investigated copy
  numbers of all protein-coding genes in \emph{Atlantogenatan} genomes to
  see if there is any correlation between the copy number of duplicates
  and changes body size along the phylogenetic tree. From an initial set
  of 18,011 protein-coding genes in hg38, we identified a median of 13,880
  genes in \emph{Atlantogenatan} genomes, of which a median of 940 genes
  are duplicated. We find that, just as body size fluctuates throughout
  \emph{Atlantogenata}, genes involved in tumor suppressor pathways are
  also duplicated throughout the phylogenetic tree. Extant species of
  elephants, however, show active transcription of both canonical and
  duplicated copies of tummor suppressors that duplicated prior to and
  during their sudden increase in body size, suggesting that the
  duplication of tumor suppressor genes facilitates the evolution of
  increased body size by compensating for the increased cancer risk.
  \end{abstract}

@@

\begin{figure}

\includegraphics[width=6in]{paper_PLOS_draft_files/figure-latex/Figure0-1} \hfill{}

\caption{Atlantogenatans with sequenced genomes, body sizes, and known lifespans [@PuttickAndThomas2015; @HAGR]}\label{fig:Figure0}
\end{figure}

@@

One of the major constraints on the evolution of large body sizes in
animals is an increased risk of developing cancer. If all cells in all
organisms have a similar risk of malignant transformation and equivalent
cancer suppression mechanisms, organism with many cells should have a
higher prevalence of cancer than organisms with fewer cells. Consistent
with this expectation there is a strong positive correlation between
body size and cancer incidence within species, for example, human cancer
incidence increases with increasing adult height {[}1,2{]} and cancer
incidence is positively correlated with body size in dogs {[}3,4{]}.
There is no correlation, however, between body size and cancer risk
between species. This lack of correlation is often referred to as
`Peto's Paradox' {[}5--7{]}. While it is clear that a resolution to
Peto's Paradox must involve the evolution of enhanced cancer protection
alongside increases in body size and lifespan, the specific genetic,
molecular, and cellular mechanisms that underlie this resistance have
proven elusive. {[}8--12{]}.

Among the challenges for discovering how animals evolved enhanced cancer
protection mechanisms is identifying lineages in which large bodied
species are nested within species with small body sizes. Afrotherian
mammals are generally small-bodied, similarly to the prediced common
ancestor of Eutherian mammals. For example, maximum adult weights are
\textasciitilde{}70g in golden moles, \textasciitilde{}120g in tenrecs,
\textasciitilde{}170g in elephant shrews, \textasciitilde{}3kg in
hyraxes, and 60kg in aardvarks {[}13{]}. However, while these extant
species are relatively small, the fossil evidence demonstrates that
their ancestral lineages reached enormous sizes. For example, while
extant hyraxes are relatively small, the extinct Titanohyrax is
estimated to have weighted up to \textasciitilde{}1300kg {[}14{]}. The
largest members of Afrotheria, too, are dwarfed by the size of their
recent ancestors: extant cows manatees are large bodied
(\textasciitilde{}322-480kg) but are relatively small compared to the
extinct Stellar's sea cow which is estimated to have weight 8000-10000kg
{[}15{]}. Similarly African (4,800kg) and Asian elephants (3,200kg) are
the largest living elephant species, but are dwarfed by the truly
gigantic extinct Proboscideans such as Deinotherium
(\textasciitilde{}132,000kg), Mammut borsoni (110,000kg), and the Asian
straight-tusked elephant (\textasciitilde{}220,000kg), the largest known
land mammal {[}16{]}. These large-bodied Afrotherian lineages are nested
within small bodied species (Fig. 1) {[}17--20{]}, indicating that
gigantism independently evolved in hyraxes, sea cows, and elephants
(Paenungulates). Thus, Paenungulates are an excellent model system in
which to explore the mechanisms that underlie the evolution of large
body sizes and augmented cancer resistance.

Although many mechanisms can potentially resolve Peto's paradox, among
the most parsimonious routes to enhanced cancer resistance is through an
increased copy number of tumor suppressors. Such an example has been
seen in the case of candidate genes such as \emph{TP53} and \emph{LIF}
{[}12,21,22{]} as well as in studies involving a limited set of
candidate genes {[}23,24{]}. As these studies focus on \emph{a priori}
gene sets, however, it remains unknown whether this is a general,
genome-wide trend in Afrotherian genomes; and whether such a general
trend is associated with the recent increases in body size -- and
therefore expected cancer risk -- in these species.

Here, we trace the evolution of body mass and gene copy number variation
in Afrotherians in order to investigate whether gene duplications are
enriched in large, long-lived species for genes involved in known tumor
suppression pathways. Our estimates of the evolution of body mass,
similarly to previous studies {[}17--20{]}, show that large body masses
evolved in a step-wise manner, with major increases in body mass in the
Pseudoungulata (17kg), Paenungulata (25kg), Tethytheria (296kg), and
Proboscidea (4,100kg) stem-lineages. Furthermore, we see that the
ancestral body size increases in Hydracoidia and Sirenia were
independent events. To study the evolution of gene copy number, we used
a genome-wide Reciprocal Best BLAT Hit (RBBH) method to identify gene
duplications in Afrotherian genomes, and used maximum likelihood
(treating copy number as a discrete trait) to infer the lineages in
which those duplications occurred. We found gene duplications in
lineages with increased body mass were enriched in functions related to
tumor suppression, including regulation of the cell cycle, DNA damage
repair, and regulation of apoptosis. These data suggest that duplication
of tumor suppressors played a role in the evolution of large, long-lived
in Afrotherians.

@@

We built a time-calibrated supertree of Eutherian mammals by combining
the time-calibrated molecular phylogeny of Bininda-Emonds \emph{et al.}
{[}25{]} with the time-calibrated total evidence Afrotherian phylogeny
from Puttick and Thomas {[}20{]}. While the Bininda-Emonds \emph{et al.}
{[}25{]} phylogeny includes 1,679 species, only 34 are Afrotherian, and
no fossil data are included. The inclusion of fossil data from extinct
species is essential to ensure that ancestral state reconstructions of
body mass are not biased by only including extant species. This can lead
to inaccurate reconstructions, for example, if lineages convergently
evolved large body masses from a small bodied ancestor. In contrast, the
total evidence Afrotherian phylogeny of Puttick and Thomas {[}20{]}
includes 77 extant species and fossil data from 39 extinct species.
Therefore we replaced the Afrotherian clade in the Bininda-Emonds
\emph{et al.} {[}25{]} phylogeny with the Afrotherian phylogeny of
Puttick and Thomas {[}20{]} using Mesquite. Next, we jointly estimated
rates of body mass evolution and reconstructed ancestral states using a
generalization of the Brownian motion model that relaxes assumptions of
neutrality and gradualism by considering increments to evolving
characters to be drawn from a heavy-tailed stable distribution (the
``Stable Model'') {[}26{]}. The stable model allows for occasional large
jumps in traits and has previously been shown to out-perform other
models of body mass evolution, including standard Brownian motion
models, Ornstein--Uhlenbeck models, early burst maximum likelihood
models, and heterogeneous multi-rate models {[}26{]}.

@@

\textbf{Reciprocal Best-Hit BLAT:} We developed a reciprocal best hit
BLAT (RBHB) pipeline to identify putative homologs and estimate gene
copy numbers across species (\textbf{Figure 1A}). The Reciprocal Best
Hit (RBH) search strategy is conceptually straightforward: 1) Given a
gene of interest \(G_A\) in a query genome \(A\), one searches a target
genome \(B\) for all possible matches to \(G_A\); 2) For each of these
hits, one then performs the reciprocal search in the original query
genome to identify the highest-scoring hit; 3) A hit in genome \(B\) is
defined as a homolog of gene \(G_A\) if and only if the original gene
\(G_A\) is the top reciprocal search hit in genome \(A\). We selected
BLAT {[}27{]} as our algorithm of choice, as this algorithm is sensitive
to highly simliar (\textgreater{}90\% identity) sequences, thus
identifying the highest-confidence homologs while minimizing many-to-one
mapping problems when searching for multiple genes. RBH performs similar
to other more complex methods of orthology prediction, and is
particularly good at identifying incomplete genes that may be fragmented
in low quality/poor assembled regions of the genome {[}28,29{]}.

\textbf{Effective Copy Number By Coverage:} In lower-quality genomes,
many genes are fragmented across multiple scaffolds, which results in
BLAT calling multiple hits when in reality there is only one gene. To
compensate for this, we came up with a novel statistic, Estimated Copy
Number by Coverage (ECNC), which averages the number of times we see
each nucleotides of a query sequence in a target genome over the total
number of nucleotides of the query sequence found overall in each target
genome (Supplementary Figure 1). This allows us to correct for genes
that have been fragmented across incomplete genomes, while also taking
into account missing sequences from the human query in the target
genome. Mathematically, this can be written as:\\
\[ ECNC = \frac{\sum_{n=1}^{l} C_n}{\sum_{n=1}^{l} bool(C_n)}\] where
\(n\) is a given nucleotide in the query, \(l\) is the total length of
the query, \(C_n\) is the number of instances that \(n\) is present
within a reciprocal best hit, and \(bool(C_n)\) is 1 if \(C_n > 0\) or 0
if \(C_n = 0\).

\textbf{RecSearch Pipeline:} We created a custom Python pipeline for
automating RBHB searches between a single reference genome and multiple
target genomes using a list of query sequences from the reference
genome. For the query sequences in our search, we used the hg38 Proteome
provided by UniProt {[}30{]}, which is a comprehensive set of protein
sequences curated from a combination of predicted and validated protein
sequences generated by the UniProt Consortium. In order to refine our
search, we omitted protein sequences originating from long, noncoding
RNA loci (e.g.~LINC genes); poorly-studied genes from predicted open
reading frames (C-ORFs); and sequences with highly repetitive sequences
such as zinc fingers, protocadherins, and transposon-containing genes,
as these were prone to high levels of false positive hits.

After filtering out problematic protein queries (see ``Query gene
inclusion criteria''), we then used our pipeline (Figure 1A) to search
for all copies of our 18011 query genes in publicly available
Afrotherian genomes, including African savannah elephant
(\emph{Loxodonta africana}: loxAfr3, loxAfr4, loxAfrC), African forest
elephant (\emph{Loxodonta cyclotis}: loxCycF), Asian Elephant
(\emph{Elephas maximus}: eleMaxD), Woolly Mammoth (\emph{Mammuthus
primigenius}: mamPriV), Colombian mammoth (\emph{Mammuthus columbi}:
mamColU), American mastodon (\emph{Mammut americanum}: mamAmeI), Rock
Hyrax (\emph{Procavia capensis}: proCap1, proCap2, proCap2\_HiC), West
Indian Manatee (\emph{Trichechus manatus latirostris}: triManLat1,
triManLat1\_HiC), Aardvark (\emph{Orycteropus afer}: oryAfe1,
oryAfe1\_HiC), Lesser Hedgehog Tenrec (\emph{Echinops telfairi}:
echTel2), Nine-banded armadillo (\emph{Dasypus novemcinctus}: dasNov3),
Hoffman's two-toed sloth (\emph{Choloepus hoffmannii}: choHof1, choHof2,
choHof2\_HiC), Cape golden mole (\emph{Chrysochloris asiatica}:
chrAsi1), and Cape elephant shrew (\emph{Elephantulus edwardii}:
eleEdw1). For many of these species, we covered multiple assemblies in
order to test the effects of assembly size and quality on our hits.

\textbf{Query gene inclusion criteria:} To assemble our query list, we
first removed all unnamed genes from UP000005640. Next, we excluded
genes from downstream analyses for which assignment of homology was
uncertain, including uncharacterized ORFs (991), LOC (63), HLA genes
(402), replication dependent histones (72), odorant receptors (499),
ribosomal proteins (410), zinc finger transcription factors (1983),
viral and repetitive-element-associated proteins (82) and any protein
described as either ``Uncharacterized,'' ``Putative,'' or ``Fragment''
by UniProt in UP000005640 (30724), leaving us with a final set of 37582
query protein sequences, corresponding to 18011 genes.

\textbf{Duplication gene inclusion criteria:} In order to condense
transcript-level hits into single gene loci, and to resolve many-to-one
genome mappings, we removed exons where transcripts from different genes
overlapped, and merged overlapping transcripts of the same gene into a
single gene locus call. The resulting gene-level copy number table was
then combined with the maximum ECNC values observed for each gene in
order to call gene duplications. We called a gene duplicated if its copy
number was two or more, and if the maximum ECNC value of all the gene
transcripts searched was 1.5 or greater; previous studies have shown
that incomplete duplications can encode functional genes {[}12,22{]},
therefore partial gene duplications were included provided they passed
additional inclusion criteria. The ECNC cut off of 1.5 was selected
empirically, as this value minimized the number of false positives seen
in a test set of genes and genomes. The results of our initial search
are summarized in Figure 1B. Overall, we identified 13880 genes across
all species, or 77.1\% of our starting query genes.

\textbf{Genome Quality Assessment using CEGMA:} In order to determine
the effect of genome quality on our results, we used the gVolante
webserver and CEGMA to assess the quality and completeness of the genome
{[}31,32{]}. CEGMA was run using the default settings for mammals
(``Cut-off length for sequence statistics and composition'' = 1;``CEGMA
max intron length'' = 100000; ``CEGMA gene flanks'' = 10000, ``Selected
reference gene set'' = CVG). For each genome, we generated a correlation
matrix using the aforementioned genome quality scores, and either the
mean Copy Number or mean ECNC for all hits in the genome.

\hypertarget{evidence-for-functionality-of-identified-genes}{%
\subsection{Evidence for Functionality of Identified
Genes}\label{evidence-for-functionality-of-identified-genes}}

@@

\textbf{Transcriptome Assembly:} For the African Savana Elephant, Asian
Elephant, West Indian Manatee, and Nine-Banded Armadillo, we generated
\emph{de novo} transcriptomes using publically-available RNA-sequencing
data from NCBI SRA (Supplementary Table 1). We mapped reads to all
genomes available for each species, and assembled transcripts using
HISAT2 and StringTie, respectively {[}33--35{]}. RNA-sequencing data was
not available for Cape Golden Mole, Cape Elephant Shrew, Rock Hyrax,
Aardvark, or the Lesser Hedgehog Tenrec.

>>>>

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\usepackage{longtable}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Pervasive duplication of tumor suppressors in Afrotherians preceded evolution of large bodies and reduced cancer risk in elephants},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

\setcounter{secnumdepth}{5}
% Pandoc toggle for numbering sections (defaults to be off)

% Pandoc header
\usepackage{longtable}
\usepackage{booktabs}
\usepackage{geometry}
\usepackage{pdflscape}
\usepackage{float}
\let\origfigure\figure
\let\endorigfigure\endfigure
\renewenvironment{figure}[1][2] {
    \expandafter\origfigure\expandafter[H]
} {
    \endorigfigure
}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

@@

  \title{Pervasive duplication of tumor suppressors in Afrotherians preceded evolution of large bodies and reduced cancer risk in elephants}
    \author[University of Chicago]{Juan Manuel Vazquez}

@@

  \begin{abstract}
  The risk of developing cancer is correlated with body size and lifespan within
  species. Between species, there is no correlation between body size or
  lifespan, and cancer, suggesting that large, long-lived species have evolved
  enhanced cancer protection mechanisms. Elephants and their relatives
  (Proboscideans) are a particularly interesting lineage in which to explore
  the mechanisms underlying the evolution of augmented cancer resistance because
  they evolved large bodies recently within a clade of smaller bodied species
  (Atlantogenatans). Here we explore the contribution of tumor suppressor
  duplication to body size and cancer risk in Atlantogenatans. Unexpectedly, we
  found that tumor suppressor duplication was pervasive in Atlantogenatan
  genomes rather than restricted to Proboscideans. Proboscideans, however, have
  duplicates in unique pathways that may underlie some aspects of their
  remarkable anti-cancer cell biology. These data suggest that duplication of
  tumor suppressor genes facilitated the evolution of increased body size by
  compensating for decreasing intrinsic cancer risk.
  \end{abstract}

@@

\newcommand{\beginsupplement}{%
        \setcounter{table}{0}
        \renewcommand{\thetable}{S.\thechapter.\arabic{table}}%
        \setcounter{figure}{0}
        \renewcommand{\thefigure}{S.\thechapter.\arabic{figure}}%
     }

@@

Among the constraints on the evolution of large bodies and long life-spans in animals is an increased risk of developing cancer. If all cells in all organisms have a similar risk of malignant transformation and equivalent cancer suppression mechanisms, then organisms with many cells should have a higher prevalence of cancer than organisms with fewer cells, particularly because large and small animals have similar cell sizes (\emph{1}). Consistent with this expectation there is a strong positive correlation between body size and cancer incidence within species, for example, cancer incidence increases with increasing adult height in humans (\emph{2}, \emph{3}) and overall body size in dogs, cats, and cattle (\emph{4}--\emph{6}). There is no correlation, however, between body size and cancer risk between species; this lack of correlation is often referred to as `Peto's Paradox' (\emph{7}--\emph{9}). Indeed, cancer prevalence is relatively stable at \textasciitilde5\% across species with diverse body sizes ranging from the minuscule 51g grass mouse to the gargantuan 4800kg African elephant (\emph{10}--\emph{12}). While the ultimate resolution to Peto's Paradox is obvious, large bodied and long-lived species evolved enhanced cancer protection mechanisms, identifying and characterizing the proximate genetic, molecular, and cellular mechanisms that underlie the evolution of augmented cancer protection has proven difficult (\emph{13}--\emph{17}).

One of the challenges for discovering how animals evolved enhanced cancer protection mechanisms is identifying lineages in which large bodied species are nested within species with small body sizes. \emph{Afrotherian} mammals are generally small-bodied, but also include the largest extant land mammals. For example, maximum adult weights are \textasciitilde70g in golden moles, \textasciitilde120g in tenrecs, \textasciitilde170g in elephant shrews, \textasciitilde3kg in hyraxes, and \textasciitilde60kg in aardvarks (\emph{18}). In contrast, while extant hyraxes are relatively small, the extinct \emph{Titanohyrax} is estimated to have weighed \textasciitilde1300kg (\emph{19}). The largest members of \emph{Afrotheria}, too, are dwarfed by the size of their recent ancestors: extant sea cows such as manatees are large bodied (\textasciitilde322-480kg) but are relatively small compared to the extinct Stellar's sea cow which is estimated to have weight \textasciitilde8000-10000kg (\emph{20}). Similarly African (4,800kg) and Asian elephants (3,200kg) are the large, but are dwarfed by the truly gigantic extinct Proboscideans such as \emph{Deinotherium} (\textasciitilde132,000kg), \emph{Mammut borsoni} (110,000kg), and the straight-tusked elephant (\textasciitilde220,000kg), the largest known land mammal (\emph{21}). Remarkably, these large-bodied \emph{Afrotherian} lineages are nested deeply within small bodied species (\textbf{Figure \ref{fig:Figure-intro}}) (\emph{22}--\emph{25}), indicating that gigantism independently evolved in hyraxes, sea cows, and elephants (\emph{Paenungulata}). Thus, \emph{Paenungulates} are an excellent model system in which to explore the mechanisms that underlie the evolution of large body sizes and augmented cancer resistance.

Many mechanisms have been suggested to resolve Peto's paradox, including a decrease in the copy number of oncogenes, an increase in the copy number of tumor suppressor genes (\emph{7}, \emph{8}, \emph{26}), reduced metabolic rates, reduced retroviral activity and load (\emph{27}), and selection for `cheater' tumors that parasitize the growth of other tumors (\emph{28}), among many others. Among the most parsimonious routes to enhanced cancer resistance may be through an increased copy number of tumor suppressors. For example, transgenic mice with additional copies of TP53 have reduced cancer rates and extended lifespans but not increased cancer (\emph{29}), respectively, suggesting that changes in the copy number of tumor suppressors can effect cancer rates. Indeed, candidate genes studies have found that the elephant genome encodes duplicate tumor suppressors such as TP53 and LIF (\emph{10}, \emph{17}, \emph{30}) as well as other genes with putative tumor suppressive functions (\emph{31}, \emph{32}). These studies, however, focused on a priori candidate genes, thus it is unclear whether duplication of tumor suppressor genes is a general phenomenon in the elephant lineage.

Here we trace the evolution of body mass, cancer risk, and gene copy number variation across Afrotherian genomes, including multiple living and extinct Proboscideans, to investigate whether duplication of tumor suppressors occurred coincident with the evolution of large body sizes. Our estimates of the evolution of body mass across Afrotheria show that large body masses evolved in a step-wise manner, similar to previous studies (\emph{22}--\emph{25}) and coincident with dramatic reductions in intrinsic cancer risk. To explore whether duplication of tumor suppressors occurred coincident with the evolution of large body sizes, we used a genome-wide Reciprocal Best BLAT Hit (RBBH) strategy to identify gene duplications, and used maximum likelihood to infer the lineages in which those duplications occurred. Unexpectedly, we found that duplication of tumor suppressor genes was common in Afrotherians, both large and small, but that gene duplications in the Proboscidean lineage were enriched in pathways that may explain some of the unique cancer protection mechanisms observed iin elephant cells. These data suggest that duplication of tumor suppressor genes is pervasive in Afrotherians and proceeded the evolution of species with very large body sizes.

@@

We first built a time-calibrated supertree of Eutherian mammals by combining the time-calibrated molecular phylogeny of Bininda-Emonds et al.~(\emph{33}) with the time-calibrated total evidence Afrotherian phylogeny from Puttick and Thomas (\emph{25}). While the Bininda-Emonds et al.~(\emph{33}) phylogeny includes 1,679 species, only 34 are Afrotherian, and no fossil data are included. The inclusion of fossil data from extinct species is essential to ensure that ancestral state reconstructions of body mass are not biased by only including extant species. This can lead to inaccurate reconstructions, for example, if lineages convergently evolved large body masses from a small bodied ancestor. In contrast, the total evidence Afrotherian phylogeny of Puttick and Thomas (\emph{25}) includes 77 extant species and fossil data from 39 extinct species. Therefore we replaced the Afrotherian clade in the Bininda-Emonds et al.~(\emph{33}) phylogeny with the Afrotherian phylogeny of Puttick and Thomas (\emph{25}) using Mesquite. Next, we jointly estimated rates of body mass evolution and reconstructed ancestral states using a generalization of the Brownian motion model that relaxes assumptions of neutrality and gradualism by considering increments to evolving characters to be drawn from a heavy-tailed stable distribution (the ``Stable Model'') implemented in StableTraits (\emph{34}). The stable model allows for large jumps in traits and has previously been shown to out-perform other models of body mass evolution, including standard Brownian motion models, Ornstein--Uhlenbeck models, early burst maximum likelihood models, and heterogeneous multi-rate models (\emph{34}).

@@

\emph{Reciprocal Best-Hit BLAT:} We developed a reciprocal best hit BLAT (RBHB) pipeline to identify putative homologs and estimate gene copy number across species. The Reciprocal Best Hit (RBH) search strategy is conceptually straightforward: 1) Given a gene of interest \(G_A\) in a query genome \(A\), one searches a target genome \(B\) for all possible matches to \(G_A\); 2) For each of these hits, one then performs the reciprocal search in the original query genome to identify the highest-scoring hit; 3) A hit in genome \(B\) is defined as a homolog of gene \(G_A\) if and only if the original gene \(G_A\) is the top reciprocal search hit in genome \(A\). We selected BLAT (\emph{35}) as our algorithm of choice, as this algorithm is sensitive to highly simliar (\textgreater90\% identity) sequences, thus identifying the highest-confidence homologs while minimizing many-to-one mapping problems when searching for multiple genes. RBH performs similar to other more complex methods of orthology prediction, and is particularly good at identifying incomplete genes that may be fragmented in low quality/poor assembled regions of the genome (\emph{36}, \emph{37}).

\emph{Effective Copy Number By Coverage:} In low-quality genomes, many genes are fragmented across multiple scaffolds, which results in BLA(S)T-like methods calling multiple hits when in reality there is only one gene. To compensate for this, developed a novel statistic, Estimated Copy Number by Coverage (ECNC), which averages the number of times we hit each nucleotide of a query sequence in a target genome over the total number of nucleotides of the query sequence found overall in each target genome (\textbf{Supplemental Figure \ref{fig:Figure-ECNC}}). This allows us to correct for genes that have been fragmented across incomplete genomes, while also taking into account missing sequences from the human query in the target genome. Mathematically, this can be written as:

\[ ECNC = \frac{\sum_{n=1}^{l} C_n}{\sum_{n=1}^{l} bool(C_n)}\]

where \(n\) is a given nucleotide in the query, \(l\) is the total length of the query, \(C_n\) is the number of instances that \(n\) is present within a reciprocal best hit, and \(bool(C_n)\) is 1 if \(C_n > 0\) or 0 if \(C_n = 0\).

\emph{RecSearch Pipeline:} We created a custom Python pipeline for automating RBHB searches between a single reference genome and multiple target genomes using a list of query sequences from the reference genome. For the query sequences in our search, we used the hg38 UniProt proteome (\emph{38}), which is a comprehensive set of protein sequences curated from a combination of predicted and validated protein sequences generated by the UniProt Consortium. In order to refine our search, we omitted protein sequences originating from long, noncoding RNA loci (e.g.~LINC genes); poorly-studied genes from predicted open reading frames (C-ORFs); and sequences with highly repetitive sequences such as zinc fingers, protocadherins, and transposon-containing genes, as these were prone to high levels of false positive hits.
After filtering out problematic protein queries, we then used our pipeline to search for all copies of 18011 query genes in publicly available \emph{Afrotherian} genomes (\ref{tab:Table-Genomes}), including African savannah elephant (\emph{Loxodonta africana}: loxAfr3, loxAfr4, loxAfrC), African forest elephant (\emph{Loxodonta cyclotis}: loxCycF), Asian Elephant (\emph{Elephas maximus}: eleMaxD), Woolly Mammoth (\emph{Mammuthus primigenius}: mamPriV), Colombian mammoth (\emph{Mammuthus columbi}: mamColU), American mastodon (\emph{Mammut americanum}: mamAmeI), Rock Hyrax (\emph{Procavia capensis}: proCap1, proCap2, proCap2\emph{HiC), West Indian Manatee (}Trichechus manatus latirostris\emph{: triManLat1, triManLat1}HiC), Aardvark (\emph{Orycteropus afer}: oryAfe1, oryAfe1\emph{HiC), Lesser Hedgehog Tenrec (}Echinops telfairi\emph{: echTel2), Nine-banded armadillo (}Dasypus novemcinctus\emph{: dasNov3), Hoffman's two-toed sloth (}Choloepus hoffmannii\emph{: choHof1, choHof2, choHof2}HiC), Cape golden mole (\emph{Chrysochloris asiatica}: chrAsi1), and Cape elephant shrew (\emph{Elephantulus edwardii}: eleEdw1).

\emph{Query gene inclusion criteria:} To assemble our query list, we first removed all unnamed genes from UP000005640. Next, we excluded genes from downstream analyses for which assignment of homology was uncertain, including uncharacterized ORFs (991 genes), LOC (63 genes), HLA genes (402 genes), replication dependent histones (72 genes), odorant receptors (499 genes), ribosomal proteins (410 genes), zinc finger transcription factors (1983 genes), viral and repetitive-element-associated proteins (82 genes) and any protein described as either ``Uncharacterized,'' ``Putative,'' or ``Fragment'' by UniProt in UP000005640 (30724 genes), leaving us with a final set of 37,582 query protein isoforms, corresponding to 18,011 genes.

\emph{Duplication gene inclusion criteria:} In order to condense transcript-level hits into single gene loci, and to resolve many-to-one genome mappings, we removed exons where transcripts from different genes overlapped, and merged overlapping transcripts of the same gene into a single gene locus call. The resulting gene-level copy number table was then combined with the maximum ECNC values observed for each gene in order to call gene duplications. We called a gene duplicated if its copy number was two or more, and if the maximum ECNC value of all the gene transcripts searched was 1.5 or greater; previous studies have shown that incomplete duplications can encode functional genes (\emph{17}, \emph{30}), therefore partial gene duplications were included provided they passed additional inclusion criteria (see below). The ECNC cut-off of 1.5 was selected empirically, as this value minimized the number of false positives seen in a test set of genes and genomes. The results of our initial search are summarized in \textbf{Figure @ref(fig: Figure-ML-UpSet-WC)A}. Overall, we identified 13880 genes across all species, or 77.1\% of our starting query genes.

\emph{Genome Quality Assessment using CEGMA:} In order to determine the effect of genome quality on our results, we used the gVolante webserver and CEGMA to assess the quality and completeness of the genome (\emph{39}, \emph{40}). CEGMA was run using the default settings for mammals (``Cut-off length for sequence statistics and composition'' = 1;``CEGMA max intron length'' = 100000; ``CEGMA gene flanks'' = 10000, ``Selected reference gene set'' = CVG). For each genome, we generated a correlation matrix using the aforementioned genome quality scores, and either the mean Copy Number or mean ECNC for all hits in the genome. We observed that the percentage of duplicated genes in non-\emph{Pseudoungulatan} genomes was higher (12.94\% to 23.66\%) than \emph{Pseudoungulatan} genomes (3.26\% to 7.80\%). Mean Copy Number, mean ECNC, and mean CN (the lesser of Copy Number and ECNC per gene) moderately or strongly correlated with genomic quality, such as LD50, the number of scaffolds, and contigs with a length above either 100K or 1M (\textbf{Supplemental Figure \ref{fig:Figure-GenomeCor}}). The Afrosoricidians had the greatest correlation between poor genome quality and high gene duplication rates including larger numbers of private duplications. The correlations between genome quality metric and number of gene duplications was particularly high for Cape golden mole (\emph{Chrysochloris asiatica}: chrAsi1) and Cape elephant shrew (\emph{Elephantulus edwardii}: eleEdw1), therefore we excluded these species from downstream pathway enrichment analyses.

\hypertarget{evidence-for-functionality-of-gene-duplicates}{%
\subsection{Evidence for Functionality of Gene Duplicates}\label{evidence-for-functionality-of-gene-duplicates}}

@@

\emph{Transcriptome Assembly}: For the African Savana Elephant, Asian Elephant, West Indian Manatee, and Nine-Banded Armadillo, we generated de novo transcriptomes using publically-available RNA-sequencing data from NCBI SRA (Supplemental Table \ref{tab:Table-SRAData}). We mapped reads to all genomes available for each species, and assembled transcripts using HISAT2 and StringTie, respectively (\emph{41}--\emph{43}). RNA-sequencing data was not available for Cape Golden Mole, Cape Elephant Shrew, Rock Hyrax, Aardvark, or the Lesser Hedgehog Tenrec.

\emph{Gene Prediction}: We downloaded tracks for genes predicted using GenScan for all the genomes available via UCSC Genome Browser: African savannah elephant (loxAfr3), Rock Hyrax (proCap1), West Indian Manatee (triManLat1), Aardvark (oryAfe1), Lesser Hedgehog Tenrec (echTel2), Nine-banded armadillo (dasNov3), Hoffman's Two-Toed Sloth (choHof1), Cape golden mole (chrAsi1), and Cape Elephant Shrew (eleEdw1); gene prediction tracks for higher-quality assemblies were not available.
